---
title: "Lab 12(HW)"
author: "Ivan Kish(PID:A17262923)"
date: "2026-02-13"
output: pdf
---


```{r}
url <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"


genes <- read.table(url, header = TRUE)


head(genes)
```
>Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
genotype_counts <- table(genes$geno)
print(genotype_counts)

median_expression <- tapply(genes$exp, genes$geno, median)
print(median_expression)
```
>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?


```{r}
library(ggplot2)

ggplot(genes) +
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE, outlier.shape = NA) + 
  geom_jitter(width = 0.2, alpha = 0.4) +          
  theme_bw() +                                   
  labs(title = "Expression by Genotype at rs8067378",
       x = "Genotype",
       y = "Expression Level") +
  theme(legend.position = "none")                  
```

Looking at he box plot we can see that homozygous A|A individuals show a  higher expression of the ORMDL3 gene with the heterozygous A|G and homozygous G|G having increasingly lower expression levels showing that depending on the content of the SNP the gene will have higher or lower levels of expression.

